resevol: an R package for spatially explicit models of pesticide resistance

### 2023-11-30 notebook/ms/advanced_techniques.Rmd
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
subtitle: Advanced techniques (Supporting Information 2)
author: |
  | A. Bradley Duthie&#xb9; &#179;, Rosie Mangan&#xb9;, Chintamani Rose McKeon&#xb9;,
  | Matthew C. Tinsley&#xb9;, and Luc F. Bussi&egrave;re&#xb2;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    fig_caption: yes
    toc: no
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
biblio-style: apalike
---

@@

To save time, a genome `mg` produced from the code above has been saved into the resevol R package as `advanced_techniques_eg.rda`, which we can load.
Note that running `mine_gmatrix` can be a time-consuming process, and will often require multiple attempts to get the above parameter settings to produce a desired network.
See @Hamblin2013 for useful advice on parameter value selection for the evolutionary algorithm.
Time-sensitive parameter values specific to the resevol package include `indivs` and `layers`.
Increasing values for both of these parameters can increase run time.

@@

We can compare the above covariance structure with the one that we specified in `gmt`. 
The stress of `mg` (i.e., mean squared deviation of `mg` elements from `gmt` elements) is `r mean((mg[[6]] - gmt)^2)`. 
But this is only based on one population of `indivs = 2000` initialised pests. 
We can use the `stress_test` function to see what the distribution of stress is over 1000 such initialised populations of 2000 individuals (Figure 1).

@@

What is important is that the matrix elements only include natural numbers from 1 to the total number of farms (not skipping any numbers). 
In the case of the above, numbers 1-7 all appear on the landscape, so the matrix will work.
We can visualise the matrix more clearly by looking at it as an image (Figure 2).

@@

In Figure 2, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
This is a simple example, but matrices can be made large to model highly complex landscapes.
Additionally, not every colour needs to represent a 'farm' per se. 
We can represent other types of terrain such as grassland, woodland, or water with its own natural number.
The trick in this case is to have each of these other types of terrain grow a crop and use a pesticide that has no effect on pests (i.e., cannot be consumed).
If terrain types have no effect on pests, then they can effectively model land that is not used for farming.
We can show this using the matrix `land_eg`, which is also included in the resevol package.
The dimensions of the landscape modelled in `land_eg` are 128 by 128 cells, and the landscape includes 14 farms (1-14), grassland (15), forest (16), and water (17).
The code below produces an image that shows how these terrain types are distributed over the landscape (Figure 3).

@@

Because the structure of the data frame above depends on the number of crops, pesticides, and evolving traits simulated, there are no pre-set functions in the resevol package for plotting.
Here we include code for plotting population size, consumption of the two food types, two pesticides types, and evolving trait means.
Note that because crop and pesticide 3 are dummy variables (not consumed at all), we do not need to plot these.
Figure 4 shows how population density of pests changes over time.

@@

Figure 5 shows how the mean consumption of different foods and pesticides changes over time.

@@

Similarly, we can visualise how all four traits evolve over time (Figure 6).

@@

We can plot the distribution over the landscape using the code below.
The code identifies the x-locations and y-locations of each pest `i` in `pop_last_time_step_sim1`, and these landscape cells are replaced with the number 18.
A new land colour for black `#000000` is then included, and the map is reproduced as an image in R.
Figure 7 shows the same map as above, but with the numbering removed to make the pests (black points) more visible.

@@

To avoid needless repetition, the code for plots will not be included for simulation 2.
Figure 8 shows population abundance over time for simulation 2.

@@

After pesticides are applied at time step 81, pest population size decreases and does not recover.
Figure 9 shows crop and pesticide consumption over time.

@@

Figure 10 shows how pest traits evolve over time when pesticides are rotated.

@@

We can observe the spatial distribution of pests on the landscape in the final time step (Figure 11).

@@

From the output above, we can conclude that when farms rotate the pesticide that they apply between pesticide 1 and pesticide 2, it has a substantial effect on pest density and evolution.
In the absence of pesticide rotation, the pest population quickly recovers after pesticides are first applied (Figure 4).
Consequently, pesticides have very little long-term effect on the amount of crops consumed (Figure 5a,b), and while there is an initial spike in the consumption of both pesticides (Figure 5c,d), this consumption quickly declines.
Despite the positive covariance between crop consumption and pesticide consumption, and the negative covariances between consumption of crop and pesticide types, these trade-offs appear to have only a modest effect on reducing actual crop consumption and maintaining or increasing pesticide consumption (Figure 6).
Consequently, some farms experience high densities of pests that are resistant to the applied pesticides (Figure 7).

In contrast, when pesticides are rotated every 9 time steps, the pest population density drops after the onset of pesticide application and does not recover (Figure 8).
The amount of each crop consumed per pests also drops (Figure 9a, b), while the amount of pesticide 1 and 2 consumed is maintained at a much higher rate than occurred in the absence of pesticide rotation (Figure 9c, d).
Because selection against pesticide consumption was not consistent for a specific farm, and there is a trade-off between pest consumption of pesticide 1 versus 2, local adaptation to a single pesticide did not occur (Figure 10).
Overall, farms therefore did not experience high densities of pests, and pest resistance to pesticide is well-managed (Figure 11).

>>>>

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
subtitle: Advanced techniques (Supporting Information 2)
author: |
  | A. Bradley Duthie&#xb9; &#179;, Rosie Mangan&#xb9;, Chintamani Rose McKeon&#xb9;,
  | Matthew C. Tinsley&#xb9;, and Luc F. Bussi&egrave;re&#xb2;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
csl: plos-computational-biology.csl
biblio-style: acm
---

\renewcommand\thefigure{\Alph{figure}}

@@

To save time, a genome `mg` produced from the code above has been saved into the resevol R package as `advanced_techniques_eg.rda`, which we can load.
Note that running `mine_gmatrix` can be a time-consuming process, and will often require multiple attempts to get the above parameter settings to produce a desired network.
See Hamblin @Hamblin2013 for useful advice on parameter value selection for the evolutionary algorithm.
Time-sensitive parameter values specific to the resevol package include `indivs` and `layers`.
Increasing values for both of these parameters can increase run time.

@@

We can compare the above covariance structure with the one that we specified in `gmt`. 
The stress of `mg` (i.e., mean squared deviation of `mg` elements from `gmt` elements) is `r mean((mg[[6]] - gmt)^2)`. 
But this is only based on one population of `indivs = 2000` initialised pests. 
We can use the `stress_test` function to see what the distribution of stress is over 1000 such initialised populations of 2000 individuals (Figure A).

@@

What is important is that the matrix elements only include natural numbers from 1 to the total number of farms (not skipping any numbers). 
In the case of the above, numbers 1-7 all appear on the landscape, so the matrix will work.
We can visualise the matrix more clearly by looking at it as an image (Figure B).

@@

In Figure B, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
This is a simple example, but matrices can be made large to model highly complex landscapes.
Additionally, not every colour needs to represent a 'farm' per se. 
We can represent other types of terrain such as grassland, woodland, or water with its own natural number.
The trick in this case is to have each of these other types of terrain grow a crop and use a pesticide that has no effect on pests (i.e., cannot be consumed).
If terrain types have no effect on pests, then they can effectively model land that is not used for farming.
We can show this using the matrix `land_eg`, which is also included in the resevol package.
The dimensions of the landscape modelled in `land_eg` are 128 by 128 cells, and the landscape includes 14 farms (1-14), grassland (15), forest (16), and water (17).
The code below produces an image that shows how these terrain types are distributed over the landscape (Figure C).

@@

Because the structure of the data frame above depends on the number of crops, pesticides, and evolving traits simulated, there are no pre-set functions in the resevol package for plotting.
Here we include code for plotting population size, consumption of the two food types, two pesticides types, and evolving trait means.
Note that because crop and pesticide 3 are dummy variables (not consumed at all), we do not need to plot these.
Figure D shows how population density of pests changes over time.

@@

Figure E shows how the mean consumption of different foods and pesticides changes over time.

@@

Similarly, we can visualise how all four traits evolve over time (Figure F).

@@

We can plot the distribution over the landscape using the code below.
The code identifies the x-locations and y-locations of each pest `i` in `pop_last_time_step_sim1`, and these landscape cells are replaced with the number 18.
A new land colour for black `#000000` is then included, and the map is reproduced as an image in R.
Figure G shows the same map as above, but with the numbering removed to make the pests (black points) more visible.

@@

To avoid needless repetition, the code for plots will not be included for simulation 2.
Figure H shows population abundance over time for simulation 2.

@@

After pesticides are applied at time step 81, pest population size decreases and does not recover.
Figure I shows crop and pesticide consumption over time.

@@

Figure J shows how pest traits evolve over time when pesticides are rotated.

@@

We can observe the spatial distribution of pests on the landscape in the final time step (Figure K).

@@

From the output above, we can conclude that when farms rotate the pesticide that they apply between pesticide 1 and pesticide 2, it has a substantial effect on pest density and evolution.
In the absence of pesticide rotation, the pest population quickly recovers after pesticides are first applied (Figure D).
Consequently, pesticides have very little long-term effect on the amount of crops consumed (Figure Ea,b), and while there is an initial spike in the consumption of both pesticides (Figure Ec,d), this consumption quickly declines.
Despite the positive covariance between crop consumption and pesticide consumption, and the negative covariances between consumption of crop and pesticide types, these trade-offs appear to have only a modest effect on reducing actual crop consumption and maintaining or increasing pesticide consumption (Figure F).
Consequently, some farms experience high densities of pests that are resistant to the applied pesticides (Figure G).

In contrast, when pesticides are rotated every 9 time steps, the pest population density drops after the onset of pesticide application and does not recover (Figure H).
The amount of each crop consumed per pests also drops (Figure Ia, b), while the amount of pesticide 1 and 2 consumed is maintained at a much higher rate than occurred in the absence of pesticide rotation (Figure Ic, d).
Because selection against pesticide consumption was not consistent for a specific farm, and there is a trade-off between pest consumption of pesticide 1 versus 2, local adaptation to a single pesticide did not occur (Figure J).
Overall, farms therefore did not experience high densities of pests, and pest resistance to pesticide is well-managed (Figure K).

### 2023-11-30 notebook/ms/evolutionary_algorithm.Rmd
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
subtitle: The evolutionary algorithm (Supporting Information 1)
author: |
  | A. Bradley Duthie&#xb9; &#179;, Rosie Mangan&#xb9;, Chintamani Rose McKeon&#xb9;,
  | Matthew C. Tinsley&#xb9;, and Luc F. Bussi&egrave;re&#xb2;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  word_document:
    reference_docx: docx_template.docx
  bookdown::pdf_document2:
    citation_package: natbib
    fig_caption: yes
    toc: no
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
biblio-style: apalike
---

@@

The objective of the evolutionary algorithm in resevol is to find a set of network values that produce traits (red diamonds above) with the pre-specified covariance structure given allele values (green circles above) that are sampled from a standard normal distribution (i.e., mean of zero and standard deviation of one). Conceptually, the problem is simple; we need to find values for the black arrows above that produce traits that covary in the way that we want them to covary (within some acceptable margin of error). The `mine_gmatrix` function uses an evolutionary algorithm for identifying sets of values that work. An evolutionary algorithm is an algorithm that works especially well for *I know it when I see it* problems [@Hamblin2013]. @Luke2013 explains the idea behind these algorithms in more detail:

@@

In the `mine_gmatrix` function of the resevol package, an evolving population of networks like that in Figure 1 above is initialised. Parent networks produce offspring networks with some recombination and mutation, and offspring networks are selected based on how close their trait covariance matrix is to the pre-specified matrix input into the function. The algorithm is inspired by a similar algorithm within the [GMSE R package](https://confoobio.github.io/gmse/index.html) [@Duthie2018].

@@

In the code, the arrows in the above Figure 1 are represented by a set of matrices that map loci values to trait values. There are 12 loci in Figure 1, and 4 nodes in each hidden layer (blue squares). Arrow values between loci and the first hidden layer can then be represented by a matrix with 12 rows and 4 columns (i.e., row 1 holds a value for each of 4 arrows that point to the 4 hidden layer nodes). Note that the values below are initialised randomly, which is how they are initialised in the evolutionary algorithm.

@@

To then get the value of the first column of four hidden layer nodes (i.e., the first column of blue squares in Figure 1), we can use matrix multiplication.

@@

The evolutionary algorithm first initialises a population of networks, with each network having a unique set of values (i.e., black arrows in Figure 1, represented in the code by matrices explained in the [previous selction](#structures)). In each iteration of the evolutionary algorithm, with some probability, networks crossover their values with another randomly selected network. Individual values in each network then mutate with some probability. The fitness of each network is then calculated by comparing its trait covariances with those of a pre-specified covariance matrix. A tournament is then used to select the highest fitness networks, and those selected networks replace the old to comprise the new population. Iterations continue until either a maximum number of iterations is reached or a network is found that produces trait covariances sufficiently close to the pre-specified covariance matrix. Figure 2 below provides a general overview of the evolutionary algorithm.

@@

An iteration of the evolutionary algorithm begins with a crossover between the values of networks in the population. For each network in the population, a crossover event will occur in the values linking loci to the first hidden layer (see Figure 1) with a probability of `pr_cross`. And a second independent crossover event will occur in the values linking the first hidden layer to traits, also with a probability of `pr_cross`. The reason that these two crossover events are independent is due to the different dimensions of the underlying arrays (see [key data structures used](#structures) above). A matrix with `loci` rows and a number of columns that matches the number of traits holds the values linking loci to the first hidden layer. A three dimensional array with row and column numbers matching trait number, and a depth matching the number of hidden `layers` (e.g., 3 in Figure 1) holds the remaining values linking the first hidden layer to trait values.

If a crossover event occurs for a focal network, then a contiguous set of values is defined and swapped with another randomly selected network in the population. Dimensions of the contiguous set are selected from a random uniform distribution. For example, given that the network in Figure 1 would be represented by a three dimensional array with 4 rows, 4 columns, and 3 layers, three random integers from 1-4, 1-4, and 1-3 would be sampled twice, respectively, with replacement. If the values selected were 1, 3, and 2 in the first sample, then 3, 3, and 1 in the second sample, then all values from rows 1-3, column 3, and layers 1-2 would be swapped between networks. Conceptually, this is the equivalent of drawing a square around set of arrows in Figure 1 and swapping the arrow values with the values of another network.

@@

After mutation, the fitness of each network in the population is evaluated. For each network, a set of `indivs` loci vectors is created, which represents the allele values of `indivs` potential individuals in a simulated population. Elements of each loci vector are randomly sampled from a standard normal distribution. For example, in the network of Figure 1 where `loci = 12`, `loci * indivs` standard normal values would be generated in total. After these `indivs` loci vectors are initialised, values in each vector are mapped to traits using the focal network, thereby producing `indivs` sets of traits from the focal network. These `indivs` sets of traits are then used to calculate the among trait covariances and build a trait covariance matrix for the network. This trait covariance matrix is then compared to the pre-specified `gmatrix` by calculating stress as the mean squared deviation between matrix elements. Lower stress values correspond to higher network fitness.

>>>>

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
subtitle: The evolutionary algorithm (Supporting Information 1)
author: |
  | A. Bradley Duthie&#xb9; &#179;, Rosie Mangan&#xb9;, Chintamani Rose McKeon&#xb9;,
  | Matthew C. Tinsley&#xb9;, and Luc F. Bussi&egrave;re&#xb2;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
csl: plos-computational-biology.csl
biblio-style: acm
---

\renewcommand\thefigure{\Alph{figure}}

@@

The objective of the evolutionary algorithm in resevol is to find a set of network values that produce traits (red diamonds above) with the pre-specified covariance structure given allele values (green circles above) that are sampled from a standard normal distribution (i.e., mean of zero and standard deviation of one). Conceptually, the problem is simple; we need to find values for the black arrows above that produce traits that covary in the way that we want them to covary (within some acceptable margin of error). The `mine_gmatrix` function uses an evolutionary algorithm for identifying sets of values that work. An evolutionary algorithm is an algorithm that works especially well for *I know it when I see it* problems [@Hamblin2013]. Luke @Luke2013 explains the idea behind these algorithms in more detail:

@@

In the `mine_gmatrix` function of the resevol package, an evolving population of networks like that in Figure A above is initialised. Parent networks produce offspring networks with some recombination and mutation, and offspring networks are selected based on how close their trait covariance matrix is to the pre-specified matrix input into the function. The algorithm is inspired by a similar algorithm within the [GMSE R package](https://confoobio.github.io/gmse/index.html) [@Duthie2018].

@@

In the code, the arrows in the above Figure A are represented by a set of matrices that map loci values to trait values. There are 12 loci in Figure A, and 4 nodes in each hidden layer (blue squares). Arrow values between loci and the first hidden layer can then be represented by a matrix with 12 rows and 4 columns (i.e., row 1 holds a value for each of 4 arrows that point to the 4 hidden layer nodes). Note that the values below are initialised randomly, which is how they are initialised in the evolutionary algorithm.

@@

To then get the value of the first column of four hidden layer nodes (i.e., the first column of blue squares in Figure A), we can use matrix multiplication.

@@

The evolutionary algorithm first initialises a population of networks, with each network having a unique set of values (i.e., black arrows in Figure A, represented in the code by matrices explained in the [previous selction](#structures)). In each iteration of the evolutionary algorithm, with some probability, networks crossover their values with another randomly selected network. Individual values in each network then mutate with some probability. The fitness of each network is then calculated by comparing its trait covariances with those of a pre-specified covariance matrix. A tournament is then used to select the highest fitness networks, and those selected networks replace the old to comprise the new population. Iterations continue until either a maximum number of iterations is reached or a network is found that produces trait covariances sufficiently close to the pre-specified covariance matrix. Figure B below provides a general overview of the evolutionary algorithm.

@@

An iteration of the evolutionary algorithm begins with a crossover between the values of networks in the population. For each network in the population, a crossover event will occur in the values linking loci to the first hidden layer (see Figure A) with a probability of `pr_cross`. And a second independent crossover event will occur in the values linking the first hidden layer to traits, also with a probability of `pr_cross`. The reason that these two crossover events are independent is due to the different dimensions of the underlying arrays (see [key data structures used](#structures) above). A matrix with `loci` rows and a number of columns that matches the number of traits holds the values linking loci to the first hidden layer. A three dimensional array with row and column numbers matching trait number, and a depth matching the number of hidden `layers` (e.g., 3 in Figure A) holds the remaining values linking the first hidden layer to trait values.

If a crossover event occurs for a focal network, then a contiguous set of values is defined and swapped with another randomly selected network in the population. Dimensions of the contiguous set are selected from a random uniform distribution. For example, given that the network in Figure A would be represented by a three dimensional array with 4 rows, 4 columns, and 3 layers, three random integers from 1-4, 1-4, and 1-3 would be sampled twice, respectively, with replacement. If the values selected were 1, 3, and 2 in the first sample, then 3, 3, and 1 in the second sample, then all values from rows 1-3, column 3, and layers 1-2 would be swapped between networks. Conceptually, this is the equivalent of drawing a square around set of arrows in Figure A and swapping the arrow values with the values of another network.

@@

After mutation, the fitness of each network in the population is evaluated. For each network, a set of `indivs` loci vectors is created, which represents the allele values of `indivs` potential individuals in a simulated population. Elements of each loci vector are randomly sampled from a standard normal distribution. For example, in the network of Figure A where `loci = 12`, `loci * indivs` standard normal values would be generated in total. After these `indivs` loci vectors are initialised, values in each vector are mapped to traits using the focal network, thereby producing `indivs` sets of traits from the focal network. These `indivs` sets of traits are then used to calculate the among trait covariances and build a trait covariance matrix for the network. This trait covariance matrix is then compared to the pre-specified `gmatrix` by calculating stress as the mean squared deviation between matrix elements. Lower stress values correspond to higher network fitness.
